---
bibliography: references.bib
authors:
  - name: Leopold Mareis
    orcid: 0000-0002-2624-6522
    affiliation: "Technical University of Munich"
    country: Germany
  - name: Mathias Drton
    orcid: 0000-0001-5614-3025
    affiliation: "Technical University of Munich"
    country: Germany
header-includes: |
  <style>
  p {
    text-align: justify;
  }
  </style>
---

# Cost-Aware Optimized Front-Door Experimental Design

| Resource | Information |
|----|----|
| Data DOI | [10.5281/zenodo.7681811](https://doi.org/10.5281/zenodo.7681811) |
| Git + DOI | [Git](https://github.com/lmareis/Cost-Aware-Optimized-Front-Door-Experimental-Design/blob/CleaR2026/README.md) [10.5281/zenodo.18960268](https://doi.org/10.5281/zenodo.18960268) |
| Paper | [arXiv](https://arxiv.org/abs/2603.22024) |
| Paper Code | [Git](https://github.com/lmareis/Cost-Aware-Optimized-Front-Door-Experimental-Design) [10.5281/zenodo.18960268](https://zenodo.org/doi/10.5281/zenodo.18960268)|
| Short Description | Demonstration of the optimized experimental design for subsequent causal effect estimation of @Mareis_Drton_2026 on the Sachs protein-signalling dataset. |

: {.striped}

## Introduction

The Sachs dataset records expression levels provides in human T cells and is accompanied by a validated ground-truth causal DAG recovered through targeted perturbation experiments [@Sachs2005; @2208.08247]. In this notebook, we present the cost-aware experimental design framework of @Mareis_Drton_2026 to two estimation problems:

-  back-door estimation of the direct effect of ***Mek*** on ***Erk***, and

-  front-door estimation of the total causal effect of ***Mek*** on ***Akt***.

While sampling, every unit contributes a baseline observation $(x_C, x_t)$; the experimenter then decides whether to additionally measure $x_M$ and, in the front-door case, $x_r$. The optimal design assigns unit-specific sampling probabilities $(\pi_1(x_C, x_t),\, \pi_2(x_C, x_t, x_M))$: units with high information-to-cost ratios for $\beta_{Mt}$ or $\beta_{rM}$ are measured more intensively, while others are subsampled. This yields a strictly lower asymptotic estimation variance than full measurement at equal total cost.

The following application computes this optimal design and reports the resulting efficiency gains, beginning with the simpler single-stage back-door problem before extending to the two-stage front-door setting.


## Problem: Cost-Aware Causal Effect Estimation

![Causal graph (left) and two-stage sampling procedure (right) for the linear front-door model. Arrows encode direct linear effects; the double-headed arc represents unobserved confounding between $X_t$ and $X_r$. The flowchart shows that $x_M$ is measured with probability $\pi_1$ in the first stage, and $x_r$ with probability $\pi_2$ in the second stage. ](DAG.png){#fig-dag fig-align="center" width="90%"}

The linear multivariate front-door model comprises confounders $X_C$, treatment $X_t$, mediator $X_M$, and response $X_r$, related by the structural equation $X = \beta X + \varepsilon$, where $\beta$ is lower-triangular and $\varepsilon$ allows for arbitrary confounding between $\varepsilon_t$ and $\varepsilon_r$; see @fig-dag. Under the front-door condition, requiring every causal path from $X_t$ to $X_r$ to pass through $X_M$, the causal effect reduces to
$$\xi = \beta_{rM}\beta_{Mt},$$
identified via the two regressions $X_t \to X_M$ and $X_M \to X_r$, each adjustable by $X_C$.

Data collection proceeds in two stages beyond the always-observed baseline $(x_C, x_t)$ at cost $c_0$. The first stage measures $x_M$ with probability $\pi_1(x_C, x_t)$ at cost $c_1$; the second measures $x_r$ with probability $\pi_2(x_C, x_t, x_M)$ at cost $c_2$. The indicator $\Delta \in \{1, 2, \infty\}$ records the stage reached. The optimal $(\pi_1^*, \pi_2^*)$ minimizes the asymptotic variance $\mathrm{Var}_\infty(\hat\xi;\,\pi)$ over regular asymptotically linear (RAL) estimators, subject to $\mathbb{E}[c_0 + \pi_1 c_1 + \pi_1\pi_2 c_2] = b_0$. The relative efficiency
$$\frac{\mathrm{Var}_\infty(\hat\xi;\,\pi^*)}{\mathrm{Var}_\infty(\hat\xi;\,1)\cdot\mathrm{scale}},$$
where scale, or *oversampling*, equalizes total cost, certifies a gain when below 1. 

The front-door factorization $\xi = \beta_{rM}\beta_{Mt}$ decomposes into two regressions: $\beta_{Mt}$ from $(X_C, X_t, X_M)$, which is a self-contained back-door problem; and $\beta_{rM}$ from $(X_C, X_t, X_M, X_r)$, the distinctly front-door component. The optimized back-door design presented below applies to any back-door estimation setting, independent of whether a response $X_r$ is available.

::: {.callout-note title="**Key insight**. Two structural observations motivate cost-aware partial-data designs.
"}

1. **Partial sufficiency.** Some estimands do not depend on all data columns: $\beta_{Mt}$ requires only $(X_C, X_t, X_M)$ and not $X_r$, while $\beta_{rM}$ additionally requires $X_r$. Estimands with partial column support can be recovered from incompletely observed units without bias, so full measurements are not needed for every unit.

2. **Budget reallocation.** Under a fixed total budget, skipping expensive later stages reduces the per-sample cost, allowing more units to enter the study. The relative efficiency quantifies the trade-off: whether the variance reduction from additional samples outweighs the information lost from partial observations.
:::

## Sachs Dataset

The Sachs dataset [@Sachs2005] records $n = 853$ simultaneous expression levels of 11 proteins and phospholipids in human T cells under cd3cd28 (CD3/CD28 antibody) stimulation. The ground-truth causal DAG, shown in @fig-sachs, was established through a combination of observational data analysis and targeted perturbation experiments [@Sachs2005; @2208.08247]. For the back-door analysis we use ***PKA*** as confounder ($X_C$), ***Mek*** as treatment ($X_t$), and ***Erk*** as mediator ($X_M$); the front-door analysis additionally uses ***Akt*** as response ($X_r$).

![Ground-truth causal DAG for the Sachs protein-signalling network [@2208.08247]. The front-door condition holds because every causal path from ***Mek*** to ***Akt*** passes through ***Erk***.](Sachs.png){#fig-sachs fig-align="center" width="60%"}

```{r, message = FALSE, warning = FALSE}
library(knitr)
#download.file("https://zenodo.org/records/7681811/files/sachs.zip?download=1",
#              "sachs.zip", mode = "wb")
#unzip("sachs.zip", files = "Data Files/cd3cd28.csv")
cd3cd28   <- read.csv(file.path("Data Files", "cd3cd28.csv"))
Sachs_mat <- as.matrix(cd3cd28)
n         <- nrow(Sachs_mat)
```

```{r, echo = FALSE}
knitr::kable(head(cd3cd28[, c("PKA", "Mek", "Erk", "Akt")]),
             digits  = 2,
             caption = "First six observations of the four variables used in the analysis (cd3cd28 condition).")
```

## Optimized Design for Back-Door Estimation $\beta_{Mt}$

Cost-aware back-door estimation of $\beta_{Mt}$ relies only on $(X_C, X_t, X_M)$. The sampling decision is: observe $(x_C, x_t)$ at cost $c_0$, then measure $x_M$ with probability $\pi_1(x_C, x_t)$ at additional cost $c_1$.

We apply this methodology with $X_C = \text{PKA}$, $X_t = \text{Mek}$, $X_M = \text{Erk}$, and $c_0 = 2$, $c_1 = 1$. In the Sachs ground-truth DAG, ***PKA*** blocks all back-door paths from ***Mek*** to ***Erk***, satisfying the back-door adjustment criterion.

The companion repository (see resource table) provides implementations for `estimate_parameters`, `compute_moments`, `compute_oif_variance`, `compute_expected_cost`, and `compute_optimal_pi`. The analysis estimates parameters from the observed data, evaluates the asymptotic variance under full measurement, and scans over oversampling levels to compute the relative efficiency.

```{r, include = FALSE}
download.file(
  "https://raw.githubusercontent.com/lmareis/Cost-Aware-Optimized-Front-Door-Experimental-Design/CleaR2026/Back_Door/function_back_door.R",
  "functions_back_door.R", mode = "wb")
writeLines(gsub("X_B", "X_C", readLines("functions_back_door.R")), "functions_back_door.R")
```

```{r, message = FALSE, warning = FALSE}
source("functions_back_door.R")
```

```{r, message = FALSE, warning = FALSE}
set.seed(1234)
data_bd <- data.frame(
  C   = rep(Inf, n),                         # Delta
  X_t = Sachs_mat[, 2],                      # Mek
  X_M = I(Sachs_mat[, 6, drop = FALSE]),     # Erk
  X_C = I(Sachs_mat[, 8, drop = FALSE])      # PKA
)

pi1_fun_bd <- function(XCt) rep(1, nrow(XCt))
c1_fun_bd  <- function(XCt) rep(1, nrow(XCt))

estimates_bd    <- estimate_parameters(data_bd, pi1_fun_bd)
mom_bd          <- compute_moments(data_bd, estimates_bd, pi1_fun_bd)
avar_vanilla_bd <- compute_oif_variance(data_bd, estimates_bd, pi1_fun_bd,
                                        pi1 = rep(1, n))
cost_vanilla_bd <- compute_expected_cost(rep(1, n), c0 = 2,
                                         c1_fun_bd(data_bd[, c("X_C", "X_t")]),
                                         mom_bd$Weight_2inf)

scales_bd   <- seq(1.05, 1.25, by = 0.05)
rel_effs_bd <- vapply(scales_bd, function(s) {
  opt      <- compute_optimal_pi(data_bd, estimates_bd,
                                 c0 = 2, c1_fun = c1_fun_bd,
                                 b0 = cost_vanilla_bd / s,
                                 pi1_0_fun = pi1_fun_bd, n_sub = 10000)
  avar_opt <- compute_oif_variance(data_bd, estimates_bd, pi1_fun_bd,
                                   pi1 = opt$pi1_star)
  avar_opt / avar_vanilla_bd / s
}, numeric(1))

scale_bd   <- 1.15
rel_eff_bd <- rel_effs_bd[3]
```

```{r, echo = FALSE}
knitr::kable(
  data.frame(Scale = scales_bd, rel_eff = round(rel_effs_bd, 4)),
  col.names = c("Scale", "Relative Efficiency"),
  align     = "cc",
  caption   = "Back-door relative efficiency for estimating the direct effect of Mek on Erk, across oversampling levels."
)
```

The table confirms that the partial sampling design achieves a variance reduction at every oversampling level considered. At $15%$ oversampling, it attains a relative efficiency of `r round(rel_eff_bd, 4)`, a `r round((1 - rel_eff_bd) * 100, 1)`% gain over uniform full measurement at equal total budget.

## Optimized Design for Front-Door Estimation of $\xi$

The back-door analysis estimated $\beta_{Mt}$, the direct effect of ***Mek*** on ***Erk***, using only $(X_C, X_t, X_M)$. The front-door setting extends this to the total causal effect $\xi = \beta_{rM}\beta_{Mt}$ of ***Mek*** on ***Akt***, which additionally requires measuring the response $X_r = \text{Akt}$. Since ***PKA*** has a direct edge to ***Akt*** in the Sachs DAG, the pair (***Mek***, ***Akt***) is confounded and the front-door path through ***Erk*** is the identification strategy; see @fig-sachs.

The two-stage design introduces a second propensity $\pi_2(x_C, x_t, x_M)$ governing whether ***Akt*** is measured after ***Erk***. The joint optimization of $(\pi_1^*, \pi_2^*)$ balances the leverage of treatment residuals against mediator residuals under the combined budget constraint, with $X_C = \text{PKA}$, $X_t = \text{Mek}$, $X_M = \text{Erk}$, $X_r = \text{Akt}$, and custom costs $c_0 = 2$, $c_1 = c_2 = 1$.

```{r, include = FALSE}
download.file(
  "https://raw.githubusercontent.com/lmareis/Cost-Aware-Optimized-Front-Door-Experimental-Design/CleaR2026/functions.R",
  "functions.R", mode = "wb")
writeLines(gsub("X_B", "X_C", readLines("functions.R")), "functions.R")
```

```{r, message = FALSE, warning = FALSE}
source("functions.R")
```

```{r, message = FALSE, warning = FALSE}
# Column indices: 2=Mek, 6=Erk, 7=Akt, 8=PKA
pi1_fun <- function(XCt)  rep(1, nrow(XCt))   # baseline: full measurement
pi2_fun <- function(XCtM) rep(1, nrow(XCtM))
c1_fun  <- function(XCt)  rep(1, nrow(XCt))   # unit cost per stage
c2_fun  <- function(XCtM) rep(1, nrow(XCtM))

set.seed(1234)
data_fd <- data.frame(
  C   = rep(Inf, n),                         # Delta
  X_t = Sachs_mat[, 2],                      # Mek
  X_M = I(Sachs_mat[, 6, drop = FALSE]),     # Erk
  X_r = Sachs_mat[, 7],                      # Akt
  X_C = I(Sachs_mat[, 8, drop = FALSE])      # PKA
)

# Estimate β_tC, β_Mt, β_MB, β_rM via nested weighted regression
estimates_fd <- estimate_parameters(data_fd, pi1_fun, pi2_fun)
beta_Mt_hat  <- as.numeric(estimates_fd$beta_Mt)
beta_rM_hat  <- as.numeric(estimates_fd$beta_rM)
```

```{r, echo = FALSE}
cat(sprintf("beta_Mt (Mek -> Erk): %.4f  (1-unit increase in Mek raises Erk by %.4f units)",
            beta_Mt_hat, beta_Mt_hat))
cat(sprintf("beta_rM (Erk -> Akt): %.4f  (1-unit increase in Erk raises Akt by %.4f units)",
            beta_rM_hat, beta_rM_hat))
cat(sprintf("Causal effect:  xi = beta_rM * beta_Mt = %.4f * %.4f = %.4f",
            beta_rM_hat, beta_Mt_hat, beta_rM_hat * beta_Mt_hat))
```

```{r, message = FALSE, warning = FALSE}
# Asymptotic variance under full measurement (pi = 1)
mom_fd          <- compute_moments(data_fd, estimates_fd, pi1_fun, pi2_fun)
avar_vanilla_fd <- compute_oif_variance(data_fd, estimates_fd, pi1_fun, pi2_fun,
                                        pi1 = rep(1, n), pi2 = rep(1, n))
cost_vanilla_fd <- compute_expected_cost(rep(1, n), rep(1, n), c0 = 2,
                                         c1_fun(data_fd[, c("X_C", "X_t")]),
                                         c2_fun(data_fd[, c("X_C", "X_t", "X_M")]),
                                         mom_fd$Weight_2inf)

# Optimal propensity (pi_1*, pi_2*) at 54% oversampling budget
scale_fd <- 1.54
opt_fd   <- compute_optimal_pi(data_fd, estimates_fd,
                               c0 = 2, c1_fun = c1_fun, c2_fun = c2_fun,
                               b0 = cost_vanilla_fd / scale_fd,
                               pi1_0_fun = pi1_fun, pi2_0_fun = pi2_fun,
                               n_sub = 10000)

# Asymptotic variance under optimal design
avar_opt_fd <- compute_oif_variance(data_fd, estimates_fd, pi1_fun, pi2_fun,
                                    pi1 = opt_fd$pi12_star[, 1],
                                    pi2 = opt_fd$pi12_star[, 2])
rel_eff_fd  <- avar_opt_fd / avar_vanilla_fd / scale_fd
```

```{r, echo = FALSE}
cat(sprintf("Front-door | scale = %.2f | relative efficiency = %.4f\n",
            scale_fd, rel_eff_fd))
```

The relative efficiency of `r round(rel_eff_fd, 4)` at 54% oversampling corresponds to a `r round((1 - rel_eff_fd) * 100, 1)`% variance reduction. The two-stage propensity achieves this by exploiting both the treatment residual (informative for $\hat\beta_{Mt}$) and the mediator residual (informative for $\hat\beta_{rM}$) to direct measurements selectively across the two stages.

## Summary

Both designs achieve strict efficiency gains at the reported oversampling levels. The larger front-door gain reflects the additional design flexibility available when both $\beta_{Mt}$ and $\beta_{rM}$ are targeted through the two-stage propensity.

```{r, echo = FALSE}
knitr::kable(
  data.frame(
    Problem   = c("Back-door", "Front-door"),
    Objective = c("Direct effect of ***Mek*** on ***Erk***",
                  "Causal effect of ***Mek*** on ***Akt***"),
    Scale     = c(scale_bd, scale_fd),
    rel_eff   = round(c(rel_eff_bd, rel_eff_fd), 4)
  ),
  col.names = c("Problem", "Objective", "Scale", "Relative Efficiency"),
  align     = "llcc",
  caption   = "Relative efficiency of the optimal cost-aware design versus full measurement at equal total budget. A relative efficiency below 1 certifies a variance reduction at equal total cost."
)
```

## Citing this Notebook {.unnumbered}

Please cite @Mareis_Drton_2026. When using the Sachs dataset, please cite @Sachs2005.

## Additional Information {.unnumbered}

**License Information:** Please follow the above DOIs for license information of data and code.

**Contact:** If you have suggestions, feel free to [create an issue](https://github.com/lmareis/Cost-Aware-Optimized-Front-Door-Experimental-Design/issues) or [contact the authors](mailto:leopold.mareis@tum.de?subject=GMCI%20Notebook%20Front-Door%20Design%20Question).

```{latex, include = FALSE}
\begin{tikzpicture}[scale = 0.6, >=Stealth,
  node style/.style={scale = 1, minimum height=0.5cm, font=\small}]
% Nodes
\node[node style] (B) at (2,2) {$\textrm{PKA}$};
\node[node style] (t) at (0,0) {$\textrm{Mek}$};
\node[node style, black!70,  fill=olive!20] (M) at (2,0) {$\textrm{Erk}$};
\node[node style, black!45, fill=olive!10] (r) at (4,0) {$\textrm{Akt}$};

% PKA edges – muted blue to distinguish from front-door path
\draw[->, teal!70!black] (B) -- (t);
\draw[->, teal!70!black] (B) -- (M);
\draw[->, teal!70!black] (B) -- (r);
\draw[->] (t) -- (M);
\draw[->] (M) -- (r);
\draw[<->, red!60!black, dashed] (t) to [out=-45,in=-135] (r);
\end{tikzpicture}
```

